Product graphics

### 2009-11-28 product-graphics.tex
Fleshing out more details.

% Aims of this paper:
%
%  * emphasise the connection between area products and factorising 
%    probabilities
% 
%  * pull all area plots on a common footing - provide a grammar for 
%    categorical graphics

@@

display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.

The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003},  equal-bin-size plots, and fluctuation diagrams. Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data. Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.

@@

@@

To illustrate these ideas, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.

@@

@@

Constraints on partitioning

\begin{itemize}
  \item Area should be proportional to weight
  \item Containment / Non-overlapping
\end{itemize}

We will add one more constraint to reduce the space of possible

\begin{itemize}
  \item Rectangular partitions
\end{itemize}

@@

\begin{itemize}
  \item {\bf bar}: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers.

@@

Plot templates combine the simple primitives into well-known graphics:

\begin{itemize}
  \item {\bf Stacked} barchart: 1 bar + $n-1$ spines in opposite direction.
  \item {\bf Nested} barchart: $n$ bars.
  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot.
  \item {\bf Double-decker} plot: $n-1$ spines + 1 spine in opposite direction.

  \item {\bf Fluctuation} diagram: 
\end{itemize}

\subsection{Violation of constraints}

\subsubsection{Area not proportional to weight}

@@

\subsubsection{Overlapping rectangles}

The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.  In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.

Example

\subsubsection{Non-rectangular partitions}

Pie charts fall out naturally, as bars in polar coordinates, with angle and radius instead of height and width. And to ensure that counts stay proportional to areas, the square-root is taken of the y-axis. This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk. Figure~\ref{fig:polar} shows some examples of product plots in polar coordinates.

@@

    \includegraphics[width=0.25\linewidth]{hb-vb-polar}%
    \includegraphics[width=0.25\linewidth]{hb-vs-polar}%
    \includegraphics[width=0.25\linewidth]{hs-hs-polar}%
    \includegraphics[width=0.25\linewidth]{hs-vs-polar}
  \caption{}
  \label{fig:polar}
\end{figure}

@@

\section{Margining}
\label{sec:margining}

Mathematically defining margining operation.

\section{Connection with log-linear models}

@@

Example dataset

@@

>>>>

% Aims of this paper:
%
%  * emphasise the connection between area graphics and factorising 
%    probabilities
% 
%  * pull all area plots on a common footing - provide a grammar for 
%    categorical graphics

@@

display categorical data with a polygon, typically a rectangle, for each combination of the factors of interest, with the area proportional to the number of observations in that combination.  Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.

@@

It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, but each category of z sums to one.

Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.

By convention we will write the top most level of the product plot on the left, and lower levels to the right.  While $f(x, y, z)$ and $f(z, y, z)$ display the same counts, their interpretation is different because of which marginal tables you see.

@@

\section{Data}
\label{sec:data}

Two ways of representing data:

\begin{itemize}
  \item Multi-dimensional contingency table
  \item Data frame, with column of counts
\end{itemize}

The two are equivalent, but it's usually easier to work with the data frame version, because variables are explicit.  Also deals better with sparsity and is common input to modelling algorithms.

Define the conditioning operator which makes the distribution uniform.

In this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.

@@

For recursive layout, need to be able to tile plane with space, and tile shape with shape.

@@

To simplify the problem, we'll initially work with three constraints on partitioning:

\begin{itemize}
  \item Area should be proportional to weight
  \item Containment / Non-overlapping
  \item Rectangular partitions
\end{itemize}

Section~\ref{sec:relax} discusses what happens when some of these constraints are relaxed.

\begin{itemize}
\end{itemize}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.25\linewidth]{part-vbar-vr}%
    \includegraphics[width=0.25\linewidth]{part-hbar-vr}%
    \includegraphics[width=0.25\linewidth]{part-vspline-vr}%
    \includegraphics[width=0.25\linewidth]{part-hspline-vr}

  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: vertical bars, horizontal bars, vertical splines and horizontal splines. }
  \label{fig:part-1d}
\end{figure}

\begin{itemize}
  \item {\bf bar}: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers, but are not space filling.  They occupy sum(values / max(values)) of the total area.

@@

The sequence of partitioning choices defines the different named area plots:

\begin{itemize}
  \item {\bf Stacked} barchart: 1 bar + $n-1$ spines in opposite direction.
  \item {\bf Nested} barchart: $n$ bars.
  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot. \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
  \item {\bf Double-decker} plot: $n-1$ spines + 1 spine in opposite direction.

  \item {\bf Fluctuation} diagram: fluct, then spline. 
  \item {\bf Equal bin size} plot: a fluctuation diagram, when the 2 variables have uniform distribution.
\end{itemize}

Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data. Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.

\section{Relaxing constraints}
\label{sec:relax}

\subsection{Area not proportional to weight}

@@

\subsection{Overlapping rectangles}

The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.  In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
This makes it easier to see all the levels of the hierarchy, not just the lowest level.

Example.

\subsection{Non-rectangular partitions}

Pie charts fall out naturally, as bars in polar coordinates, with angle and radius instead of height and width. And to ensure that counts stay proportional to areas, the square-root is taken of the y-axis. This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk. Figure~\ref{fig:polar} illustrates a few product plots in polar coordinates.  Variables are happiness (top level in four categories) and sex (nested, two levels).

@@

    \includegraphics[width=0.25\linewidth]{hb-vb-polar}%
    \includegraphics[width=0.25\linewidth]{hb-vs-polar}%
    \includegraphics[width=0.25\linewidth]{hs-hs-polar}%
    \includegraphics[width=0.25\linewidth]{hs-vs-polar}
  \caption{(Top) Area graphics in Cartesian coordinates.  (Bottom) Area graphics in polar coordinates.  From left to right: horizontal bar nested inside horizontal bar, vertical spline nested inside horizontal bar (stacked bar chart), horizontal spline nested in horizontal spline, horizontal spline nested in vertical spline (mosaic plot).}
  \label{fig:polar}
\end{figure}

@@

% \section{Connection with log-linear models}

@@

Example dataset - Doug Bates teachers/schools?

@@

\section{Data analysis} % (fold)
\label{sec:data_analysis}

% section data_analysis (end)
